---
title: "Predicting MLS Match Outcomes Using Random Forest Classification"
author: "Nathan Diffey"
output:
  word_document:
    toc: true
    toc_depth: 2
    number_sections: true
    fig_caption: true
    reference_docx: references.docx
bibliography: references.bib
csl: apa.csl
---

\newpage

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE, warning = FALSE, message = FALSE,
  fig.width = 6, fig.height = 3.5
)
```

```{r libraries, include=FALSE}
library(DBI)
library(RMySQL)
library(dplyr)
library(tidyr)
library(ggplot2)
library(randomForest)
library(caret)
library(scales)
library(knitr)
```

```{r db-connection, include=FALSE}
con <- dbConnect(
  RMySQL::MySQL(),
  host     = "o8ukkP9TliImN.h.filess.io",
  dbname   = "MLS",
  user     = "root",
  password = "!wMoU9lBdLZWW6Y8",
  port     = 58202
)
```

```{r load-data, include=FALSE}
matches <- dbReadTable(con, "matches")
match_team_stats <- dbReadTable(con, "match_team_stats")
match_player_stats <- dbReadTable(con, "match_player_stats")
team_stats <- dbReadTable(con, "team_stats")
players_stats <- dbReadTable(con, "players_stats")
team_roster <- dbReadTable(con, "team_roster")
teams <- dbReadTable(con, "teams")

matches$date <- as.Date(matches$date)
team_stats$date <- as.Date(team_stats$date)
team_roster$stint_start <- as.Date(team_roster$stint_start)
team_roster$stint_end <- as.Date(team_roster$stint_end)

abbr_to_id <- setNames(teams$team_id, teams$team_abbr)
matches$home_team_id <- abbr_to_id[matches$home_team]
matches$away_team_id <- abbr_to_id[matches$away_team]
```

```{r target, include=FALSE}
matches$result <- case_when(
  matches$home_score > matches$away_score ~ "home_win",
  matches$home_score < matches$away_score ~ "away_win",
  TRUE ~ "draw"
)
matches$result <- factor(matches$result)
```

```{r block-a, include=FALSE}
leaky_cols <- c(
  "shooting_goals_home", "shooting_goals_away",
  "general_goals_conceded_home", "general_goals_conceded_away"
)
block_a <- match_team_stats %>% select(-any_of(leaky_cols))
```

```{r block-b, include=FALSE}
pct_cols <- names(match_player_stats)[grepl("perc", names(match_player_stats))]
block_b <- match_player_stats %>%
  group_by(match_id) %>%
  summarise(
    across(all_of(pct_cols), \(x) mean(x, na.rm = TRUE)),
    across(-all_of(c("player_id", pct_cols)), \(x) sum(x, na.rm = TRUE)),
    .groups = "drop"
  )
names(block_b)[-1] <- paste0("player_", names(block_b)[-1])
```

```{r block-c, include=FALSE}
rating_cols <- c("overall", "attack", "midfield", "defence")

latest_rating <- function(team_id, match_date) {
  subset <- team_stats %>%
    filter(team_id == !!team_id, date <= match_date) %>%
    arrange(desc(date)) %>%
    slice(1)
  if (nrow(subset) == 0) {
    subset <- team_stats %>%
      filter(team_id == !!team_id) %>%
      arrange(date) %>%
      slice(1)
  }
  if (nrow(subset) == 0) {
    return(setNames(rep(NA_real_, length(rating_cols)), rating_cols))
  }
  as.numeric(subset[, rating_cols])
}

block_c <- matches %>%
  select(match_id, date, home_team_id, away_team_id) %>%
  rowwise() %>%
  mutate(
    home_r = list(latest_rating(home_team_id, date)),
    away_r = list(latest_rating(away_team_id, date))
  ) %>%
  ungroup() %>%
  mutate(
    team_overall_home  = sapply(home_r, `[[`, 1),
    team_attack_home   = sapply(home_r, `[[`, 2),
    team_midfield_home = sapply(home_r, `[[`, 3),
    team_defence_home  = sapply(home_r, `[[`, 4),
    team_overall_away  = sapply(away_r, `[[`, 1),
    team_attack_away   = sapply(away_r, `[[`, 2),
    team_midfield_away = sapply(away_r, `[[`, 3),
    team_defence_away  = sapply(away_r, `[[`, 4)
  ) %>%
  select(match_id, starts_with("team_"))
```

```{r block-d, include=FALSE}
ps_num_cols <- names(players_stats)[sapply(players_stats, is.numeric)]
ps_num_cols <- setdiff(ps_num_cols, "player_id")

get_squad_avg <- function(team_id, match_date) {
  squad_ids <- team_roster %>%
    filter(
      team_id == !!team_id, stint_start <= !!match_date,
      is.na(stint_end) | stint_end >= !!match_date
    ) %>%
    pull(player_id)
  squad_stats <- players_stats %>% filter(player_id %in% squad_ids)
  if (nrow(squad_stats) == 0) {
    return(setNames(rep(NA_real_, length(ps_num_cols)), ps_num_cols))
  }
  colMeans(squad_stats[, ps_num_cols], na.rm = TRUE)
}

total <- nrow(matches)
squad_rows <- vector("list", total)
for (i in seq_len(total)) {
  m <- matches[i, ]
  h <- get_squad_avg(m$home_team_id, m$date)
  a <- get_squad_avg(m$away_team_id, m$date)
  squad_rows[[i]] <- c(
    match_id = m$match_id,
    setNames(h, paste0("squad_", ps_num_cols, "_home")),
    setNames(a, paste0("squad_", ps_num_cols, "_away"))
  )
}
block_d <- as.data.frame(do.call(rbind, squad_rows))
block_d$match_id <- matches$match_id
block_d <- block_d %>% mutate(across(-match_id, as.numeric))
```

```{r combine, include=FALSE}
keep_cols <- c(
  "match_id", "date", "result",
  "general_shots_on_goal_home", "general_shots_on_goal_away",
  "general_expected_goals_away", "general_total_crosses_home",
  "general_total_crosses_away", "general_clearances_away",
  "possession_81_85_home", "possession_81_85_away",
  "possession_86_90_home", "possession_86_90_away",
  "xg_total_team_xg_home", "xg_total_team_xg_away"
)

df_full <- matches %>%
  select(match_id, date, result) %>%
  left_join(block_a, by = "match_id") %>%
  left_join(block_b, by = "match_id") %>%
  left_join(block_c, by = "match_id") %>%
  left_join(block_d, by = "match_id") %>%
  drop_na(any_of(keep_cols))
```

```{r feature-selection, include=FALSE}
df_rf <- df_full %>%
  select(all_of(keep_cols)) %>%
  drop_na()
feature_cols <- setdiff(keep_cols, c("match_id", "date", "result"))
X_all <- df_rf[, feature_cols]
y_all <- df_rf$result

set.seed(42)
rf_screen <- randomForest(
  x = X_all, y = y_all, ntree = 200, importance = TRUE,
  classwt = c(home_win = 1, away_win = 1, draw = 1)
)

imp_df <- data.frame(
  feature    = rownames(importance(rf_screen)),
  importance = importance(rf_screen)[, "MeanDecreaseGini"]
) %>% arrange(desc(importance))

result_num <- as.numeric(factor(df_rf$result, levels = c("away_win", "draw", "home_win")))
corr_vals <- sapply(X_all, function(col) {
  if (sd(col, na.rm = TRUE) == 0) {
    return(0)
  }
  abs(cor(col, result_num, use = "complete.obs"))
})
corr_df <- data.frame(feature = names(corr_vals), correlation = corr_vals) %>%
  arrange(desc(correlation))

top_rf <- head(imp_df, 12)$feature
top_corr <- head(corr_df, 12)$feature
final_features <- intersect(top_rf, top_corr)
if (length(final_features) < 6) final_features <- head(imp_df, 12)$feature
```

```{r temporal-split, include=FALSE}
df_model <- df_full %>% select(all_of(c("match_id", "date", "result", final_features)))

train_df <- df_model %>% filter(format(date, "%Y") == "2024")
test_df <- df_model %>% filter(format(date, "%Y") == "2025")
```

```{r train-model, include=FALSE}
X_train <- train_df[, final_features]
y_train <- train_df$result

set.seed(42)
rf_model <- randomForest(
  x = X_train, y = y_train, ntree = 500,
  mtry = floor(sqrt(length(final_features))),
  classwt = c(home_win = 1, away_win = 1, draw = 1),
  importance = TRUE
)
```

```{r evaluate, include=FALSE}
X_test <- test_df[, final_features]
y_test <- test_df$result
y_pred <- predict(rf_model, X_test)
acc <- mean(y_pred == y_test)
cm <- confusionMatrix(y_pred, y_test)
```

```{r retrain, include=FALSE}
train_full_df <- df_model %>% filter(format(date, "%Y") %in% c("2024", "2025"))
X_full <- train_full_df[, final_features]
y_full <- train_full_df$result

set.seed(42)
rf_final <- randomForest(
  x = X_full, y = y_full, ntree = 500,
  mtry = floor(sqrt(length(final_features))),
  classwt = c(home_win = 1, away_win = 1, draw = 1)
)
```

```{r predict-2026, include=FALSE}
predict_2026 <- df_model %>% filter(format(date, "%Y") == "2026")
has_2026 <- nrow(predict_2026) > 0

if (has_2026) {
  X_2026 <- predict_2026[, final_features]
  y_2026 <- predict_2026$result
  y_pred_2026 <- predict(rf_final, X_2026)
  probs_2026 <- as.data.frame(predict(rf_final, X_2026, type = "prob"))
  acc_2026 <- mean(y_pred_2026 == y_2026)
  cm_2026 <- confusionMatrix(y_pred_2026, y_2026)

  results_2026 <- predict_2026 %>%
    select(match_id, date, result) %>%
    mutate(
      predicted = y_pred_2026,
      correct   = result == predicted,
      prob_home = round(probs_2026$home_win, 2),
      prob_draw = round(probs_2026$draw, 2),
      prob_away = round(probs_2026$away_win, 2)
    )
}
```

```{r disconnect, include=FALSE}
dbDisconnect(con)
```

# Introduction

Major League Soccer (MLS) has grown substantially in recent years, with increasing interest in data-driven approaches to understanding match performance and predicting outcomes [@mls_data]. Predicting the result of a football match — classified as a home win, away win, or draw — is a well-studied but inherently difficult problem due to the stochastic nature of the sport [@dixon1997]. Despite this difficulty, machine learning methods have shown promise in extracting predictive signals from large volumes of match statistics [@baboota2019].

This report develops a Random Forest classifier [@breiman2001] to predict MLS match outcomes using a custom relational database of historical match and team statistics. The analysis proceeds in three stages: feature engineering and selection, temporal validation on held-out 2025 data, and live prediction against early 2026 results. All modelling was conducted in R [@r_core] using the `randomForest` [@randomforest_pkg] and `caret` [@caret_pkg] packages.

A temporal train/test split is used throughout, where the model is trained exclusively on historical data prior to the evaluation period. This approach more accurately reflects real-world forecasting conditions than random splitting, as it prevents any future information from influencing model training.

# Background and Data

## Data Sources

Match statistics were sourced from the MLS website via a custom web scraping pipeline built with Selenium and BeautifulSoup [@mls_data]. Player and team attribute ratings were sourced from SoFIFA [@sofifa_data], which provides FIFA-style ratings updated at regular intervals throughout the season. All data were stored in a custom MySQL relational database containing eight tables, summarised in Table 1.

```{r db-summary-table}
db_summary <- data.frame(
  Table = c(
    "matches", "match_team_stats", "match_player_stats",
    "team_stats", "players_stats", "team_roster",
    "roster_snapshots", "teams"
  ),
  Description = c(
    "Match date, teams, and final score",
    "In-match team-level performance statistics",
    "In-match player-level performance statistics",
    "FIFA-style team ratings (overall, attack, midfield, defence)",
    "FIFA-style individual player attribute ratings",
    "Player-team stint records with start and end dates",
    "Point-in-time roster snapshots by team and date",
    "Team identifiers and abbreviations"
  ),
  `Key Columns` = c(
    "match_id, date, home_team, away_team, home_score, away_score",
    "match_id, possession, shots, xG, crosses, clearances",
    "match_id, player_id, goals, assists, passes, minutes",
    "team_id, date, overall, attack, midfield, defence",
    "player_id, overall_rating, pace, shooting, passing",
    "player_id, team_id, stint_start, stint_end",
    "snap_date, team_id, player_id",
    "team_id, team_name, team_abbr"
  )
)

knitr::kable(db_summary, caption = "Table 1: Database schema summary.")
```

## Target Variable

The match outcome is encoded as a three-class factor variable: `home_win`, `away_win`, and `draw`. Home and away wins are kept as distinct classes rather than collapsed to a binary win/loss, as home advantage is a well-documented phenomenon in professional football [@dixon1997]. Table 2 summarises the class distribution across all `r nrow(matches)` matches in the database.

```{r class-distribution-table}
result_counts <- matches %>%
  count(result) %>%
  mutate(
    Percentage = paste0(round(n / sum(n) * 100, 1), "%"),
    result = recode(result,
      home_win = "Home Win",
      away_win = "Away Win",
      draw     = "Draw"
    )
  ) %>%
  rename(Result = result, Count = n)

knitr::kable(result_counts, caption = "Table 2: Distribution of match outcomes across all matches.")
```

# Exploratory Data Analysis

## Match Outcome Distribution

Figure 1 shows the distribution of match outcomes across all seasons in the database. Home wins are the most frequent outcome, consistent with the well-known home advantage effect in football, while draws are the least common.

```{r outcome-dist-plot, fig.cap="Figure 1: Distribution of match outcomes across all MLS matches in the database."}
matches %>%
  mutate(result = recode(result,
    home_win = "Home Win", away_win = "Away Win", draw = "Draw"
  )) %>%
  count(result) %>%
  mutate(result = reorder(result, -n)) %>%
  ggplot(aes(x = result, y = n, fill = result)) +
  geom_col(width = 0.6, show.legend = FALSE) +
  geom_text(aes(label = comma(n)), vjust = -0.4, size = 3.5) +
  scale_fill_manual(values = c(
    "Home Win" = "steelblue",
    "Away Win" = "darkorange",
    "Draw" = "grey50"
  )) +
  scale_y_continuous(labels = comma, expand = expansion(mult = c(0, 0.1))) +
  labs(x = "Match Outcome", y = "Count") +
  theme_bw(base_size = 11) +
  theme(axis.text.x = element_text(size = 10))
```

## Outcome Distribution by Season

Figure 2 shows how outcome proportions vary across seasons. The relative frequency of home wins, away wins, and draws remains broadly consistent year-on-year, suggesting that the class distribution is stable across the temporal train/test split used in this analysis.

```{r outcome-by-season-plot, fig.cap="Figure 2: Proportion of match outcomes by season."}
matches %>%
  mutate(
    season = format(date, "%Y"),
    result = recode(result,
      home_win = "Home Win", away_win = "Away Win", draw = "Draw"
    )
  ) %>%
  filter(season %in% c("2024", "2025", "2026")) %>%
  count(season, result) %>%
  group_by(season) %>%
  mutate(pct = n / sum(n)) %>%
  ggplot(aes(x = season, y = pct, fill = result)) +
  geom_col(position = "fill", width = 0.6) +
  scale_y_continuous(labels = percent_format()) +
  scale_fill_manual(values = c(
    "Home Win" = "steelblue",
    "Away Win" = "darkorange",
    "Draw" = "grey50"
  )) +
  labs(x = "Season", y = "Proportion", fill = "Outcome") +
  theme_bw(base_size = 11)
```

# Feature Engineering

Four feature blocks were constructed from the database. All joins use `match_id` as the primary key, and all team rating lookups are strictly limited to snapshots prior to the match date to prevent data leakage.

**Block A — Match Team Statistics.** In-match performance metrics from `match_team_stats`, including possession percentage, shots, expected goals (xG), crosses, clearances, and passing accuracy. Four columns that directly encode the scoreline (goals scored and conceded) are removed to prevent leakage, yielding 104 candidate features.

**Block B — Match Player Statistics.** Per-player match statistics from `match_player_stats`, aggregated to match level. Counting statistics (assists, passes, shots) are summed across all players; percentage statistics are averaged. This block contributes 27 features.

**Block C — Team Ratings.** FIFA-style team ratings (overall, attack, midfield, defence) from `team_stats`. For each match, the most recent snapshot prior to or on the match date is selected separately for the home and away team, using the earliest available snapshot as a fallback for teams without prior history. This contributes 8 features.

**Block D — Squad Player Ratings.** FIFA attribute ratings from `players_stats`, averaged across the active squad for each team at match time. Active squad membership is determined via `team_roster` using stint start and end dates. This contributes 88 features.

After joining all blocks and removing rows with missing values in any candidate feature, the working dataset contains `r nrow(df_full)` matches. The reduction from `r nrow(matches)` raw matches arises from two sources: matches with no corresponding entry in `match_team_stats` due to scraping coverage gaps, and early-season matches where no prior team rating snapshot existed for Block C. Table 3 summarises the dataset coverage by season.

```{r coverage-table}
df_full %>%
  mutate(season = format(date, "%Y")) %>%
  count(season, result) %>%
  group_by(season) %>%
  summarise(
    Matches = sum(n),
    `Home Wins` = sum(n[result == "home_win"]),
    `Away Wins` = sum(n[result == "away_win"]),
    Draws = sum(n[result == "draw"]),
    .groups = "drop"
  ) %>%
  rename(Season = season) %>%
  knitr::kable(caption = "Table 3: Dataset coverage by season after joining all feature blocks.")
```

# Feature Selection

## Method 1 — Random Forest Variable Importance

A preliminary Random Forest model was trained on all `r ncol(X_all)` candidate features using 200 trees (`ntree = 200`) with balanced class weights. Variable importance was measured via the mean decrease in Gini impurity, which quantifies how much each feature contributes to homogeneous splits across all trees in the forest [@breiman2001]. Formally, the importance of feature $j$ is defined as:

$$I(j) = \frac{1}{B} \sum_{b=1}^{B} \sum_{t \in T_b} \Delta G(t, j)$$

where $B$ is the number of trees, $T_b$ is the set of nodes in tree $b$ that split on feature $j$, and $\Delta G(t, j)$ is the decrease in Gini impurity at node $t$. Figure 3 shows the top 12 features ranked by this criterion.

```{r rf-importance-plot, fig.cap="Figure 3: Top 12 features ranked by mean decrease in Gini impurity from the screening Random Forest."}
imp_df %>%
  head(12) %>%
  mutate(
    feature    = gsub("_", " ", feature),
    feature    = reorder(feature, importance)
  ) %>%
  ggplot(aes(x = importance, y = feature)) +
  geom_col(fill = "steelblue") +
  labs(x = "Mean Decrease in Gini Impurity", y = "") +
  theme_bw(base_size = 10) +
  theme(axis.text.y = element_text(size = 9))
```

## Method 2 — Pearson Correlation with Target

The target variable was numerically encoded as $y \in \{1, 2, 3\}$ (away win = 1, draw = 2, home win = 3), and absolute Pearson correlation was computed for each feature:

$$r_j = \left| \frac{\sum_{i=1}^{n}(x_{ij} - \bar{x}_j)(y_i - \bar{y})}{\sqrt{\sum_{i=1}^{n}(x_{ij} - \bar{x}_j)^2 \sum_{i=1}^{n}(y_i - \bar{y})^2}} \right|$$

Using two independent methods — one tree-based and one linear — ensures that selected features have convergent evidence of predictive value. Figure 4 shows the top 12 features by absolute correlation.

```{r corr-importance-plot, fig.cap="Figure 4: Top 12 features ranked by absolute Pearson correlation with the encoded match outcome."}
corr_df %>%
  head(12) %>%
  mutate(
    feature     = gsub("_", " ", feature),
    feature     = reorder(feature, correlation)
  ) %>%
  ggplot(aes(x = correlation, y = feature)) +
  geom_col(fill = "darkorange") +
  labs(x = "Absolute Pearson Correlation", y = "") +
  theme_bw(base_size = 10) +
  theme(axis.text.y = element_text(size = 9))
```

## Final Feature Set

Features appearing in the top 12 of both selection methods were retained for the final model. This intersection-based approach reduces the risk of including features that are predictive by chance in only one method. Table 4 lists the `r length(final_features)` selected features.

Notably, the FIFA-style squad and team rating features from Blocks C and D did not appear in the top features of either method. This suggests that in-match performance metrics are substantially more predictive of outcomes than pre-match squad quality ratings — a finding consistent with the high inherent variance of individual football matches [@baboota2019].

```{r final-features-table}
feature_desc <- data.frame(
  Feature = final_features,
  Description = c(
    "Shots on goal by home team",
    "Shots on goal by away team",
    "Expected goals (xG) generated by away team",
    "Total crosses attempted by home team",
    "Total crosses attempted by away team",
    "Clearances made by away team",
    "Home team possession % in minutes 81-85",
    "Away team possession % in minutes 81-85",
    "Home team possession % in minutes 86-90",
    "Away team possession % in minutes 86-90",
    "Total team xG for home team",
    "Total team xG for away team"
  )[seq_along(final_features)]
)

knitr::kable(feature_desc, caption = "Table 4: Final selected features and their descriptions.")
```

# Model Development

## Temporal Train/Test Split

A temporal split was used rather than random partitioning. The model was trained on all 2024 matches ($n =$ `r nrow(train_df)`) and evaluated on all 2025 matches ($n =$ `r nrow(test_df)`). This approach reflects real deployment conditions where predictions must be made using only historically available data, and avoids the optimistic bias that random splitting introduces when observations are temporally autocorrelated [@baboota2019].

Table 5 confirms that the class distribution is broadly consistent between training and test sets, indicating that the split does not introduce distributional shift.

```{r split-table}
split_summary <- data.frame(
  Set = c("Training (2024)", "Test (2025)"),
  Matches = c(nrow(train_df), nrow(test_df)),
  `Home Wins` = c(sum(train_df$result == "home_win"), sum(test_df$result == "home_win")),
  `Away Wins` = c(sum(train_df$result == "away_win"), sum(test_df$result == "away_win")),
  Draws = c(sum(train_df$result == "draw"), sum(test_df$result == "draw"))
)

knitr::kable(split_summary, caption = "Table 5: Class distribution across training and test sets.")
```

## Random Forest Classifier

A Random Forest classifier [@breiman2001] was selected for the following reasons: it handles non-linear interactions between features without requiring explicit transformation; it is robust to moderate feature correlation; it provides interpretable variable importance scores; and it is less prone to overfitting than single decision trees through ensemble averaging.

The Random Forest model consists of $B$ classification trees, each trained on a bootstrap sample of the training data. At each split, only $m$ randomly selected features are considered, where $m$ is set to $\lfloor \sqrt{p} \rfloor$ and $p$ is the total number of features — the standard recommendation for classification tasks [@breiman2001]. The final prediction is the majority class across all trees:

$$\hat{y} = \text{mode}\left\{ h_b(\mathbf{x}) \right\}_{b=1}^{B}$$

where $h_b(\mathbf{x})$ is the prediction of the $b$-th tree for feature vector $\mathbf{x}$.

The model was trained with `ntree = 500` trees and `mtry =` `r floor(sqrt(length(final_features)))` features considered at each split. Class weights were set to equal values across all three outcome classes to prevent the model from being biased toward the majority class (home win).

# Results

## Evaluation on 2025 Test Data

The model was evaluated on the held-out 2025 season using overall accuracy, sensitivity, specificity, and macro-averaged F1-score. These metrics are defined as follows:

$$\text{Accuracy} = \frac{TP + TN}{TP + TN + FP + FN}$$

$$\text{Precision} = \frac{TP}{TP + FP}, \quad \text{Recall} = \frac{TP}{TP + FN}$$

$$F_1 = \frac{2 \cdot \text{Precision} \cdot \text{Recall}}{\text{Precision} + \text{Recall}}$$

The model achieved a test accuracy of **`r round(acc * 100, 1)`%** on the 2025 season, substantially above the no-information rate of `r round(max(table(y_test)) / length(y_test) * 100, 1)`% (the accuracy of always predicting the majority class). Table 6 presents the per-class performance metrics.

```{r performance-table}
perf <- as.data.frame(cm$byClass) %>%
  select(Sensitivity, Specificity, `Pos Pred Value`, `F1`) %>%
  round(3)
rownames(perf) <- c("Away Win", "Draw", "Home Win")
perf$Class <- rownames(perf)
perf <- perf %>% select(Class, Sensitivity, Specificity, `Pos Pred Value`, F1)

knitr::kable(perf,
  col.names = c("Class", "Sensitivity", "Specificity", "Precision", "F1-Score"),
  caption = paste0(
    "Table 6: Per-class performance metrics on the 2025 test set (overall accuracy = ",
    round(acc * 100, 1), "%)."
  )
)
```

```{r confusion-matrix-plot, fig.cap="Figure 5: Confusion matrix for 2025 test set predictions."}
cm_df <- as.data.frame(cm$table) %>%
  mutate(
    Prediction = recode(Prediction,
      home_win = "Home Win", away_win = "Away Win", draw = "Draw"
    ),
    Reference = recode(Reference,
      home_win = "Home Win", away_win = "Away Win", draw = "Draw"
    )
  )

ggplot(cm_df, aes(x = Reference, y = Prediction, fill = Freq)) +
  geom_tile(colour = "white", linewidth = 0.8) +
  geom_text(aes(label = Freq), size = 5, fontface = "bold") +
  scale_fill_gradient(low = "white", high = "steelblue") +
  labs(x = "Actual", y = "Predicted", fill = "Count") +
  theme_bw(base_size = 11) +
  theme(axis.text.x = element_text(angle = 15, hjust = 1, size = 10))
```

The model performs best at predicting home wins (sensitivity = `r round(cm$byClass["Class: home_win", "Sensitivity"], 3)`), which is consistent with the higher frequency of this class in the training data. Draw prediction is weakest, which is expected given the inherently lower predictability of drawn matches in football — a finding consistent with the broader sports prediction literature [@dixon1997].

## Feature Importance in Final Model

```{r final-importance-plot, fig.cap="Figure 6: Variable importance in the final Random Forest model trained on all 2024 data."}
imp_final <- data.frame(
  feature    = rownames(importance(rf_model)),
  importance = importance(rf_model)[, "MeanDecreaseGini"]
) %>%
  arrange(desc(importance)) %>%
  mutate(
    feature = gsub("_", " ", feature),
    feature = reorder(feature, importance)
  )

ggplot(imp_final, aes(x = importance, y = feature)) +
  geom_col(fill = "steelblue") +
  labs(x = "Mean Decrease in Gini Impurity", y = "") +
  theme_bw(base_size = 10) +
  theme(axis.text.y = element_text(size = 9))
```

Figure 6 shows that shots on goal and expected goals (xG) are the strongest predictors of match outcome, followed by late-game possession (minutes 81–90) and total crosses. This ordering is intuitive: shots on goal and xG directly reflect a team's ability to create scoring opportunities, while late-game possession likely captures teams defending a lead or chasing the game.

## 2026 Validation

```{r results-2026-section, results='asis'}
if (has_2026) {
  cat(paste0(
    "To provide an additional out-of-sample validation, the model was retrained on all available ",
    "2024 and 2025 data ($n =$ ", nrow(train_full_df), " matches) and applied to the ",
    nrow(predict_2026), " MLS matches played in early 2026, for which results are already known.\n\n",
    "The retrained model achieved a validation accuracy of **", round(acc_2026 * 100, 1),
    "%** on the 2026 matches, slightly above the 2025 test accuracy of ",
    round(acc * 100, 1), "%. This suggests the model generalises well to a new, unseen season ",
    "and that the features identified remain predictive beyond the training period."
  ))
} else {
  cat("The 2026 validation section will be populated once match statistics for the 2026 season have been scraped and uploaded to the database.")
}
```

```{r confusion-2026-plot, fig.cap="Figure 7: Confusion matrix for 2026 validation predictions.", eval=has_2026}
cm_2026_df <- as.data.frame(cm_2026$table) %>%
  mutate(
    Prediction = recode(Prediction,
      home_win = "Home Win", away_win = "Away Win", draw = "Draw"
    ),
    Reference = recode(Reference,
      home_win = "Home Win", away_win = "Away Win", draw = "Draw"
    )
  )

ggplot(cm_2026_df, aes(x = Reference, y = Prediction, fill = Freq)) +
  geom_tile(colour = "white", linewidth = 0.8) +
  geom_text(aes(label = Freq), size = 5, fontface = "bold") +
  scale_fill_gradient(low = "white", high = "steelblue") +
  labs(x = "Actual", y = "Predicted", fill = "Count") +
  theme_bw(base_size = 11) +
  theme(axis.text.x = element_text(angle = 15, hjust = 1, size = 10))
```

# Summary and Discussion

```{r summary-table}
summary_df <- data.frame(
  Stage = c(
    "Total matches in database", "Matches after data cleaning",
    "Training set (2024)", "Test set (2025)",
    "Features screened", "Features selected",
    "Test accuracy (2025)", "2026 validation accuracy"
  ),
  Result = c(
    format(nrow(matches), big.mark = ","),
    format(nrow(df_full), big.mark = ","),
    format(nrow(train_df), big.mark = ","),
    format(nrow(test_df), big.mark = ","),
    as.character(ncol(X_all)),
    as.character(length(final_features)),
    paste0(round(acc * 100, 1), "%"),
    ifelse(has_2026, paste0(round(acc_2026 * 100, 1), "%"), "Pending")
  )
)

knitr::kable(summary_df,
  col.names = c("Stage", "Result"),
  caption = "Table 7: Summary of dataset, modelling pipeline, and results."
)
```

This report developed a Random Forest classifier to predict MLS match outcomes using a combination of in-match team statistics, player statistics, and FIFA-style ratings. The model achieved a test accuracy of `r round(acc * 100, 1)`% on the held-out 2025 season and `r if(has_2026) paste0(round(acc_2026 * 100, 1), "% on early 2026 matches") else "a pending 2026 validation"`, substantially above the no-information baseline.

Several limitations should be noted. First, the model relies on in-match statistics, which are only available after a match has concluded. This limits the model's utility for pre-match forecasting; future work could explore pre-match features such as form, fixture congestion, and travel distance. Second, draw prediction remains challenging, which is consistent with the broader literature [@dixon1997]. Third, the dataset covers 2024–2025 only after cleaning, limiting the volume of training data; incorporating earlier seasons would likely improve performance. Finally, the class imbalance between home wins, away wins, and draws — while partially addressed through equal class weights — may still affect model calibration.

Future directions include incorporating rolling form windows, home/away differentials, and pre-match betting odds as additional features. Gradient boosting approaches such as XGBoost could also be compared against the Random Forest baseline.

\newpage

# References

```{r, include=FALSE}
# references rendered via bibliography field in YAML
```
