---
title: "MLS Match Result Prediction — Feature Selection"
author: "Your Name"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
    highlight: tango
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

## 1. Introduction

This analysis builds a predictive model to classify MLS match outcomes as a **home win**, **away win**, or **draw**. Before training the full model, we first identify the most informative features from our database using two independent methods:

1. Random Forest variable importance (Gini impurity)
2. Absolute Pearson correlation with the encoded target variable

Features that rank highly in **both** methods are selected for the final model, providing convergent evidence of predictive value.

---

## 2. Libraries & Database Connection

```{r libraries}
library(DBI)
library(RMySQL)
library(dplyr)
library(tidyr)
library(ggplot2)
library(randomForest)
library(corrplot)
library(caret)
```

```{r db-connection}
con <- dbConnect(
  RMySQL::MySQL(),
  host     = "localhost",
  dbname   = "MLS",
  user     = "your_user",
  password = "your_password"
)
```

---

## 3. Load Data

```{r load-data}
matches            <- dbReadTable(con, "matches")
match_team_stats   <- dbReadTable(con, "match_team_stats")
match_player_stats <- dbReadTable(con, "match_player_stats")
team_stats         <- dbReadTable(con, "team_stats")
players_stats      <- dbReadTable(con, "players_stats")
team_roster        <- dbReadTable(con, "team_roster")
teams              <- dbReadTable(con, "teams")

# Parse dates
matches$date           <- as.Date(matches$date)
team_stats$date        <- as.Date(team_stats$date)
team_roster$stint_start <- as.Date(team_roster$stint_start)
team_roster$stint_end   <- as.Date(team_roster$stint_end)

# Map team abbreviation -> team_id
abbr_to_id <- setNames(teams$team_id, teams$team_abbr)
matches$home_team_id <- abbr_to_id[matches$home_team]
matches$away_team_id <- abbr_to_id[matches$away_team]

cat("Matches loaded:", nrow(matches), "\n")
```

---

## 4. Create Target Variable

We encode the match result as a three-class factor. Home and away wins are kept separate rather than collapsed to win/loss, as home advantage is a well-documented phenomenon in football research.

```{r target}
matches$result <- case_when(
  matches$home_score > matches$away_score ~ "home_win",
  matches$home_score < matches$away_score ~ "away_win",
  TRUE                                    ~ "draw"
)
matches$result <- factor(matches$result)

table(matches$result)
```

---

## 5. Feature Engineering

### 5.1 Block A — Match Team Stats

We remove four columns that directly encode the match result (goals scored/conceded) to prevent data leakage.

```{r block-a}
leaky_cols <- c("shooting_goals_home", "shooting_goals_away",
                "general_goals_conceded_home", "general_goals_conceded_away")

block_a <- match_team_stats %>%
  select(-any_of(leaky_cols))

cat("Block A features:", ncol(block_a) - 1, "\n")
```

### 5.2 Block B — Match Player Stats (aggregated per match)

Player-level stats are aggregated to match level. Counting stats (goals, assists, etc.) are summed; percentage stats are averaged.

```{r block-b}
pct_cols <- names(match_player_stats)[grepl("perc", names(match_player_stats))]

block_b <- match_player_stats %>%
  group_by(match_id) %>%
  summarise(
    across(all_of(pct_cols), mean, na.rm = TRUE),
    across(-all_of(c("match_id", "player_id", pct_cols)), sum, na.rm = TRUE),
    .groups = "drop"
  )

names(block_b)[-1] <- paste0("player_", names(block_b)[-1])

cat("Block B features:", ncol(block_b) - 1, "\n")
```

### 5.3 Block C — Team Ratings (most recent before match date)

For each match we look up the most recent FIFA-style team rating snapshot **before** the match date, separately for home and away teams. Using only pre-match data avoids leakage.

```{r block-c}
rating_cols <- c("overall", "attack", "midfield", "defence")

latest_rating <- function(team_id, match_date) {
  subset <- team_stats %>%
    filter(team_id == !!team_id, date < !!match_date) %>%
    arrange(desc(date)) %>%
    slice(1)
  if (nrow(subset) == 0) return(setNames(rep(NA_real_, length(rating_cols)), rating_cols))
  as.numeric(subset[, rating_cols])
}

cat("Joining team ratings...\n")
block_c <- matches %>%
  select(match_id, date, home_team_id, away_team_id) %>%
  rowwise() %>%
  mutate(
    home_r = list(latest_rating(home_team_id, date)),
    away_r = list(latest_rating(away_team_id, date))
  ) %>%
  ungroup() %>%
  mutate(
    team_overall_home  = sapply(home_r, `[[`, 1),
    team_attack_home   = sapply(home_r, `[[`, 2),
    team_midfield_home = sapply(home_r, `[[`, 3),
    team_defence_home  = sapply(home_r, `[[`, 4),
    team_overall_away  = sapply(away_r, `[[`, 1),
    team_attack_away   = sapply(away_r, `[[`, 2),
    team_midfield_away = sapply(away_r, `[[`, 3),
    team_defence_away  = sapply(away_r, `[[`, 4)
  ) %>%
  select(match_id, starts_with("team_"))

cat("Block C features:", ncol(block_c) - 1, "\n")
```

### 5.4 Block D — Squad Player Ratings (averaged via team roster)

For each match, active squad members are identified via `team_roster` using stint dates, and their FIFA attribute ratings from `players_stats` are averaged per team.

```{r block-d}
ps_num_cols <- names(players_stats)[sapply(players_stats, is.numeric)]
ps_num_cols <- setdiff(ps_num_cols, "player_id")

get_squad_avg <- function(team_id, match_date) {
  squad_ids <- team_roster %>%
    filter(
      team_id == !!team_id,
      stint_start <= !!match_date,
      is.na(stint_end) | stint_end >= !!match_date
    ) %>%
    pull(player_id)

  squad_stats <- players_stats %>% filter(player_id %in% squad_ids)
  if (nrow(squad_stats) == 0) return(setNames(rep(NA_real_, length(ps_num_cols)), ps_num_cols))
  colMeans(squad_stats[, ps_num_cols], na.rm = TRUE)
}

cat("Averaging squad ratings (slowest step)...\n")
total <- nrow(matches)
squad_rows <- vector("list", total)

for (i in seq_len(total)) {
  if (i %% 100 == 0 || i == total) cat("  Squad ratings:", i, "/", total, "\n")
  m <- matches[i, ]
  h <- get_squad_avg(m$home_team_id, m$date)
  a <- get_squad_avg(m$away_team_id, m$date)
  entry <- c(match_id = m$match_id,
             setNames(h, paste0("squad_", ps_num_cols, "_home")),
             setNames(a, paste0("squad_", ps_num_cols, "_away")))
  squad_rows[[i]] <- entry
}

block_d <- as.data.frame(do.call(rbind, squad_rows))
block_d$match_id <- matches$match_id
block_d <- block_d %>% mutate(across(-match_id, as.numeric))

cat("Block D features:", ncol(block_d) - 1, "\n")
```

---

## 6. Combine All Blocks

```{r combine}
df <- matches %>%
  select(match_id, result) %>%
  left_join(block_a, by = "match_id") %>%
  left_join(block_b, by = "match_id") %>%
  left_join(block_c, by = "match_id") %>%
  left_join(block_d, by = "match_id") %>%
  drop_na()

cat("Final dataset:", nrow(df), "matches,", ncol(df) - 2, "features\n")
table(df$result)
```

---

## 7. Feature Importance — Method 1: Random Forest (Gini)

A Random Forest classifier is trained on all features. Variable importance is measured via mean decrease in Gini impurity — features that consistently split the data in an informative way receive higher scores. `class.weights` is set to handle the class imbalance between home wins, away wins, and draws.

```{r rf-importance, fig.width=9, fig.height=7}
feature_cols <- setdiff(names(df), c("match_id", "result"))
X <- df[, feature_cols]
y <- df$result

set.seed(42)
rf_model <- randomForest(
  x          = X,
  y          = y,
  ntree      = 200,
  importance = TRUE,
  classwt    = c(home_win = 1, away_win = 1, draw = 1)
)

imp_df <- data.frame(
  feature    = rownames(importance(rf_model)),
  importance = importance(rf_model)[, "MeanDecreaseGini"]
) %>% arrange(desc(importance))

cat("Top 20 — Random Forest Gini importance:\n")
print(head(imp_df, 20))

# Plot
imp_df %>%
  head(20) %>%
  mutate(feature = reorder(feature, importance)) %>%
  ggplot(aes(x = importance, y = feature)) +
  geom_col(fill = "steelblue") +
  labs(title = "Feature Importance — Random Forest (Gini)",
       x = "Mean Decrease in Gini", y = "") +
  theme_minimal(base_size = 12)
```

---

## 8. Feature Importance — Method 2: Correlation with Target

The target is numerically encoded (home_win = 2, draw = 1, away_win = 0) and Pearson correlation is computed for each feature. Absolute values are used since both positive and negative correlations indicate predictive power.

```{r correlation-importance, fig.width=9, fig.height=7}
result_num <- as.numeric(factor(df$result, levels = c("away_win", "draw", "home_win")))

corr_vals <- sapply(X, function(col) {
  if (sd(col, na.rm = TRUE) == 0) return(0)
  abs(cor(col, result_num, use = "complete.obs"))
})

corr_df <- data.frame(
  feature     = names(corr_vals),
  correlation = corr_vals
) %>% arrange(desc(correlation))

cat("Top 20 — Absolute correlation with result:\n")
print(head(corr_df, 20))

# Plot
corr_df %>%
  head(20) %>%
  mutate(feature = reorder(feature, correlation)) %>%
  ggplot(aes(x = correlation, y = feature)) +
  geom_col(fill = "darkorange") +
  labs(title = "Feature Correlation with Match Result",
       x = "Absolute Pearson Correlation", y = "") +
  theme_minimal(base_size = 12)
```

---

## 9. Redundancy Check

Features with perfect or near-perfect correlation are redundant — keeping both would not add information but would increase model complexity and risk overfitting.

```{r redundancy}
pairs_to_check <- list(
  c("shooting_on_target_away",  "xg_shots_on_target_away"),
  c("shooting_on_target_away",  "general_shots_on_goal_away"),
  c("shooting_on_target_home",  "xg_shots_on_target_home"),
  c("shooting_on_target_home",  "general_shots_on_goal_home"),
  c("general_xg_conceded_home", "general_expected_goals_away"),
  c("general_xg_conceded_away", "general_expected_goals_home"),
  c("possession_81_85_home",    "possession_86_90_home"),
  c("possession_81_85_away",    "possession_86_90_away")
)

redundancy_results <- lapply(pairs_to_check, function(p) {
  data.frame(
    feature_a   = p[1],
    feature_b   = p[2],
    correlation = round(cor(df[[p[1]]], df[[p[2]]], use = "complete.obs"), 4)
  )
})

redundancy_df <- do.call(rbind, redundancy_results)
print(redundancy_df)
```

Features with correlation = 1.0 are identical — we drop the duplicates and keep one representative per group.

---

## 10. Final Feature Selection

Features that appear in the **top 20 of both methods** and are **not redundant** with another selected feature are retained.

```{r final-features}
top_rf   <- head(imp_df, 20)$feature
top_corr <- head(corr_df, 20)$feature
agreed   <- intersect(top_rf, top_corr)

# Drop redundant features — keep one from each perfectly correlated pair
drop_redundant <- c(
  "xg_shots_on_target_away",   # identical to general_shots_on_goal_away
  "shooting_on_target_away",   # identical to general_shots_on_goal_away
  "xg_shots_on_target_home",   # identical to general_shots_on_goal_home
  "shooting_on_target_home",   # identical to general_shots_on_goal_home
  "general_xg_conceded_home",  # identical to general_expected_goals_away
  "general_xg_conceded_away"   # identical to general_expected_goals_home
)

final_features <- setdiff(agreed, drop_redundant)

cat("Final selected features (", length(final_features), "):\n")
for (f in sort(final_features)) cat(" ", f, "\n")
```

---

## 11. Summary

```{r summary}
cat("Dataset size:      ", nrow(df), "matches\n")
cat("Total features:    ", length(feature_cols), "\n")
cat("Agreed (top 20):   ", length(agreed), "\n")
cat("After redundancy:  ", length(final_features), "\n\n")
cat("These features will be used in the full Random Forest model (next step).\n")
```

```{r disconnect, include=FALSE}
dbDisconnect(con)
```
