---
title: "Predicting MLS Match Outcomes Using Random Forest Classification"
author: "Nathan Diffey"
output:
  word_document:
    toc: true
    toc_depth: 2
    number_sections: true
    fig_caption: true
bibliography: references.bib
csl: apa.csl
---

\newpage

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE, warning = FALSE, message = FALSE,
  fig.width = 5.5, fig.height = 2.8
)
```

```{r libraries, include=FALSE}
library(DBI)
library(RMySQL)
library(dplyr)
library(tidyr)
library(ggplot2)
library(randomForest)
library(caret)
library(scales)
library(knitr)
library(reshape2)
```

```{r db-connection, include=FALSE}
con <- dbConnect(
  RMySQL::MySQL(),
  host     = "o8ukkP9TliImN.h.filess.io",
  dbname   = "MLS",
  user     = "root",
  password = "!wMoU9lBdLZWW6Y8",
  port     = 58202
)
```

```{r load-data, include=FALSE}
matches <- dbReadTable(con, "matches")
match_team_stats <- dbReadTable(con, "match_team_stats")
match_player_stats <- dbReadTable(con, "match_player_stats")
team_stats <- dbReadTable(con, "team_stats")
players_stats <- dbReadTable(con, "players_stats")
team_roster <- dbReadTable(con, "team_roster")
teams <- dbReadTable(con, "teams")

matches$date <- as.Date(matches$date)
team_stats$date <- as.Date(team_stats$date)
team_roster$stint_start <- as.Date(team_roster$stint_start)
team_roster$stint_end <- as.Date(team_roster$stint_end)

abbr_to_id <- setNames(teams$team_id, teams$team_abbr)
matches$home_team_id <- abbr_to_id[matches$home_team]
matches$away_team_id <- abbr_to_id[matches$away_team]
```

```{r target, include=FALSE}
matches$result <- case_when(
  matches$home_score > matches$away_score ~ "home_win",
  matches$home_score < matches$away_score ~ "away_win",
  TRUE ~ "draw"
)
matches$result <- factor(matches$result)
```

```{r block-a, include=FALSE}
leaky_cols <- c(
  "shooting_goals_home", "shooting_goals_away",
  "general_goals_conceded_home", "general_goals_conceded_away"
)
block_a <- match_team_stats %>% select(-any_of(leaky_cols))
```

```{r block-b, include=FALSE}
pct_cols <- names(match_player_stats)[grepl("perc", names(match_player_stats))]
block_b <- match_player_stats %>%
  group_by(match_id) %>%
  summarise(
    across(all_of(pct_cols), \(x) mean(x, na.rm = TRUE)),
    across(-all_of(c("player_id", pct_cols)), \(x) sum(x, na.rm = TRUE)),
    .groups = "drop"
  )
names(block_b)[-1] <- paste0("player_", names(block_b)[-1])
```

```{r block-c, include=FALSE}
rating_cols <- c("overall", "attack", "midfield", "defence")

latest_rating <- function(team_id, match_date) {
  subset <- team_stats %>%
    filter(team_id == !!team_id, date <= match_date) %>%
    arrange(desc(date)) %>%
    slice(1)
  if (nrow(subset) == 0) {
    subset <- team_stats %>%
      filter(team_id == !!team_id) %>%
      arrange(date) %>%
      slice(1)
  }
  if (nrow(subset) == 0) {
    return(setNames(rep(NA_real_, length(rating_cols)), rating_cols))
  }
  as.numeric(subset[, rating_cols])
}

block_c <- matches %>%
  select(match_id, date, home_team_id, away_team_id) %>%
  rowwise() %>%
  mutate(
    home_r = list(latest_rating(home_team_id, date)),
    away_r = list(latest_rating(away_team_id, date))
  ) %>%
  ungroup() %>%
  mutate(
    team_overall_home  = sapply(home_r, `[[`, 1),
    team_attack_home   = sapply(home_r, `[[`, 2),
    team_midfield_home = sapply(home_r, `[[`, 3),
    team_defence_home  = sapply(home_r, `[[`, 4),
    team_overall_away  = sapply(away_r, `[[`, 1),
    team_attack_away   = sapply(away_r, `[[`, 2),
    team_midfield_away = sapply(away_r, `[[`, 3),
    team_defence_away  = sapply(away_r, `[[`, 4)
  ) %>%
  select(match_id, starts_with("team_"))
```

```{r block-d, include=FALSE}
ps_num_cols <- names(players_stats)[sapply(players_stats, is.numeric)]
ps_num_cols <- setdiff(ps_num_cols, "player_id")

get_squad_avg <- function(team_id, match_date) {
  squad_ids <- team_roster %>%
    filter(
      team_id == !!team_id, stint_start <= !!match_date,
      is.na(stint_end) | stint_end >= !!match_date
    ) %>%
    pull(player_id)
  squad_stats <- players_stats %>% filter(player_id %in% squad_ids)
  if (nrow(squad_stats) == 0) {
    return(setNames(rep(NA_real_, length(ps_num_cols)), ps_num_cols))
  }
  colMeans(squad_stats[, ps_num_cols], na.rm = TRUE)
}

total <- nrow(matches)
squad_rows <- vector("list", total)
for (i in seq_len(total)) {
  m <- matches[i, ]
  h <- get_squad_avg(m$home_team_id, m$date)
  a <- get_squad_avg(m$away_team_id, m$date)
  squad_rows[[i]] <- c(
    match_id = m$match_id,
    setNames(h, paste0("squad_", ps_num_cols, "_home")),
    setNames(a, paste0("squad_", ps_num_cols, "_away"))
  )
}
block_d <- as.data.frame(do.call(rbind, squad_rows))
block_d$match_id <- matches$match_id
block_d <- block_d %>% mutate(across(-match_id, as.numeric))
```

```{r combine, include=FALSE}
# Join all blocks into one wide dataset
df_joined <- matches %>%
  select(match_id, date, result) %>%
  left_join(block_a, by = "match_id") %>%
  left_join(block_b, by = "match_id") %>%
  left_join(block_c, by = "match_id") %>%
  left_join(block_d, by = "match_id")
```

```{r redundancy-check, include=FALSE}
# Identify all home/away pairs and compute their mean sum
# to determine which are mathematically complementary (sum ~ 100)
# or percentage stats that should be converted to differentials

home_cols <- names(df_joined)[grepl("_home$", names(df_joined))]
away_cols <- gsub("_home$", "_away", home_cols)
valid <- away_cols %in% names(df_joined)

home_cols <- home_cols[valid]
away_cols <- away_cols[valid]

pair_sums <- data.frame(
  home_col = home_cols,
  away_col = away_cols,
  mean_sum = sapply(seq_along(home_cols), function(i) {
    h <- df_joined[[home_cols[i]]]
    a <- df_joined[[away_cols[i]]]
    mean(h + a, na.rm = TRUE)
  })
) %>% arrange(desc(mean_sum))

# Decision rule:
# Convert to differential (home - away) if mean sum >= 100.
# This captures:
#   (a) Possession bands and general_possession_pct which sum to exactly 100%
#   (b) Passing/overall percentage stats (~168) which measure the same
#       construct from both perspectives
#   (c) Squad aggregate totals and team ratings which sum above 100
# Keep as independent features if mean sum < 100
# (count stats like shots, crosses, fouls, clearances carry independent info)

# line 1: differential pairs
diff_pairs <- pair_sums %>%
  filter(mean_sum >= 100)

# line 2: independent pairs
independent_pairs <- pair_sums %>%
  filter(mean_sum < 100) %>%
  filter(!is.nan(mean_sum))
```

```{r apply-differentials, include=FALSE}
# Apply differential transformation to qualifying pairs
# Differential = home - away (positive = home advantage)
df_transformed <- df_joined

for (i in seq_len(nrow(diff_pairs))) {
  h_col <- diff_pairs$home_col[i]
  a_col <- diff_pairs$away_col[i]
  # Name the new column by stripping _home suffix and adding _diff
  new_col <- gsub("_home$", "_diff", h_col)
  df_transformed[[new_col]] <- df_transformed[[h_col]] - df_transformed[[a_col]]
}

# Remove original home/away columns that were converted to differentials
cols_to_remove <- c(diff_pairs$home_col, diff_pairs$away_col)
df_transformed <- df_transformed %>% select(-any_of(cols_to_remove))

# Also remove match_id duplicates from joins and non-numeric junk
# Remove entirely-NA columns first to prevent drop_na() removing all rows
df_transformed <- df_transformed %>%
  select(where(~ !all(is.na(.))))

df_full <- df_transformed %>%
  select(match_id, date, result, where(is.numeric)) %>%
  drop_na()
```

```{r feature-screening, include=FALSE}
feature_cols <- setdiff(names(df_full), c("match_id", "date", "result"))
X_all <- df_full[, feature_cols]
y_all <- df_full$result

set.seed(42)
rf_screen <- randomForest(
  x = X_all, y = y_all, ntree = 200, importance = TRUE,
  classwt = c(home_win = 1, away_win = 1, draw = 1)
)

imp_df <- data.frame(
  feature    = rownames(importance(rf_screen)),
  importance = importance(rf_screen)[, "MeanDecreaseGini"]
) %>% arrange(desc(importance))

result_num <- as.numeric(factor(df_full$result, levels = c("away_win", "draw", "home_win")))
corr_vals <- sapply(X_all, function(col) {
  if (sd(col, na.rm = TRUE) == 0) {
    return(0)
  }
  abs(cor(col, result_num, use = "complete.obs"))
})
corr_df <- data.frame(feature = names(corr_vals), correlation = corr_vals) %>%
  arrange(desc(correlation))

top_rf <- head(imp_df, 15)$feature
top_corr <- head(corr_df, 15)$feature
final_features <- intersect(top_rf, top_corr)
if (length(final_features) < 4) final_features <- head(imp_df, 15)$feature
```

```{r temporal-split, include=FALSE}
df_model <- df_full %>% select(all_of(c("match_id", "date", "result", final_features)))

train_df <- df_model %>% filter(format(date, "%Y") == "2024")
test_df <- df_model %>% filter(format(date, "%Y") == "2025")
```

```{r train-model, include=FALSE}
X_train <- train_df[, final_features]
y_train <- train_df$result

set.seed(42)
rf_model <- randomForest(
  x = X_train, y = y_train, ntree = 500,
  mtry = floor(sqrt(length(final_features))),
  classwt = c(home_win = 1, away_win = 1, draw = 1),
  importance = TRUE
)
```

```{r evaluate, include=FALSE}
X_test <- test_df[, final_features]
y_test <- test_df$result
y_pred <- predict(rf_model, X_test)
acc <- mean(y_pred == y_test)
cm <- confusionMatrix(y_pred, y_test)
```

```{r retrain, include=FALSE}
train_full_df <- df_model %>% filter(format(date, "%Y") %in% c("2024", "2025"))
X_full <- train_full_df[, final_features]
y_full <- train_full_df$result

set.seed(42)
rf_final <- randomForest(
  x = X_full, y = y_full, ntree = 500,
  mtry = floor(sqrt(length(final_features))),
  classwt = c(home_win = 1, away_win = 1, draw = 1)
)
```

```{r predict-2026, include=FALSE}
predict_2026 <- df_model %>% filter(format(date, "%Y") == "2026")
has_2026 <- nrow(predict_2026) > 0

if (has_2026) {
  X_2026 <- predict_2026[, final_features]
  y_2026 <- predict_2026$result
  y_pred_2026 <- predict(rf_final, X_2026)
  probs_2026 <- as.data.frame(predict(rf_final, X_2026, type = "prob"))
  acc_2026 <- mean(y_pred_2026 == y_2026)
  cm_2026 <- confusionMatrix(y_pred_2026, y_2026)

  results_2026 <- predict_2026 %>%
    select(match_id, date, result) %>%
    mutate(
      predicted = y_pred_2026,
      correct   = result == predicted,
      prob_home = round(probs_2026$home_win, 2),
      prob_draw = round(probs_2026$draw, 2),
      prob_away = round(probs_2026$away_win, 2)
    )
}
```

```{r disconnect, include=FALSE}
dbDisconnect(con)
```

# Introduction

Major League Soccer (MLS) has grown substantially in recent years, with increasing interest in data-driven approaches to predicting match outcomes [@mls_data]. Predicting the result of a football match — classified as a home win, away win, or draw — is inherently difficult due to the stochastic nature of the sport [@dixon1997], though machine learning methods have shown promise in extracting predictive signals from match statistics [@baboota2019].

This report develops a Random Forest classifier [@breiman2001] to predict MLS match outcomes using a custom relational database of historical match and team statistics, covering feature engineering, temporal validation on 2025 data, and live prediction against early 2026 results. All modelling was conducted in R [@r_core] using the `randomForest` [@randomforest_pkg] and `caret` [@caret_pkg] packages. A temporal train/test split is used throughout to reflect real-world forecasting conditions.

# Background and Data

## Data Sources

Match statistics were sourced from the MLS website via a custom web scraping pipeline built with Selenium and BeautifulSoup [@mls_data]. Player and team attribute ratings were sourced from SoFIFA [@sofifa_data], which provides FIFA-style ratings updated at regular intervals throughout the season. All data were stored in a custom MySQL relational database containing eight tables, summarised in Table 1.

```{r db-summary-table}
db_summary <- data.frame(
  Table = c(
    "matches", "match_team_stats", "match_player_stats",
    "team_stats", "players_stats", "team_roster",
    "roster_snapshots", "teams"
  ),
  Description = c(
    "Match date, teams, and final score",
    "In-match team-level performance statistics",
    "In-match player-level performance statistics",
    "FIFA-style team ratings (overall, attack, midfield, defence)",
    "FIFA-style individual player attribute ratings",
    "Player-team stint records with start and end dates",
    "Point-in-time roster snapshots by team and date",
    "Team identifiers and abbreviations"
  )
)

knitr::kable(db_summary, caption = "Table 1: Database schema summary.")
```

## Target Variable

The match outcome is encoded as a three-class factor variable: `home_win`, `away_win`, and `draw`. Home and away wins are kept as distinct classes rather than collapsed to a binary win/loss, as home advantage is a well-documented phenomenon in professional football [@dixon1997]. Table 2 summarises the class distribution across all `r nrow(matches)` matches in the database.

```{r class-distribution-table}
result_counts <- matches %>%
  count(result) %>%
  mutate(
    Percentage = paste0(round(n / sum(n) * 100, 1), "%"),
    result = recode(result,
      home_win = "Home Win",
      away_win = "Away Win",
      draw     = "Draw"
    )
  ) %>%
  rename(Result = result, Count = n)

knitr::kable(result_counts, caption = "Table 2: Distribution of match outcomes across all matches.")
```

# Exploratory Data Analysis

## Match Outcome Distribution

Figure 1 shows the distribution of match outcomes across all seasons in the database. Home wins are the most frequent outcome, consistent with the well-known home advantage effect in football, while draws are the least common.

```{r outcome-dist-plot, fig.cap="Figure 1: Distribution of match outcomes across all MLS matches in the database."}
matches %>%
  mutate(result = recode(result,
    home_win = "Home Win", away_win = "Away Win", draw = "Draw"
  )) %>%
  count(result) %>%
  mutate(result = reorder(result, -n)) %>%
  ggplot(aes(x = result, y = n, fill = result)) +
  geom_col(width = 0.6, show.legend = FALSE) +
  geom_text(aes(label = comma(n)), vjust = -0.4, size = 3.5) +
  scale_fill_manual(values = c(
    "Home Win" = "steelblue",
    "Away Win" = "darkorange",
    "Draw" = "grey50"
  )) +
  scale_y_continuous(labels = comma, expand = expansion(mult = c(0, 0.1))) +
  labs(x = "Match Outcome", y = "Count") +
  theme_bw(base_size = 11) +
  theme(axis.text.x = element_text(size = 10))
```

## Outcome Distribution by Season

The class distribution remains broadly consistent across 2024, 2025, and 2026, confirming that the temporal train/test split does not introduce distributional shift.

# Feature Engineering

Four feature blocks were constructed from the database. All joins use `match_id` as the primary key, and team rating lookups are strictly limited to snapshots prior to the match date to prevent data leakage.

**Block A — Match Team Statistics.** In-match metrics from `match_team_stats` including possession, shots, xG, crosses, and clearances. Four columns encoding the scoreline are removed to prevent leakage, yielding 104 candidate features.

**Block B — Match Player Statistics.** Per-player stats from `match_player_stats` aggregated to match level — counting stats summed, percentage stats averaged. Contributes 27 features.

**Block C — Team Ratings.** FIFA-style team ratings from `team_stats` (overall, attack, midfield, defence), using the most recent snapshot prior to each match date. Contributes 8 features.

**Block D — Squad Player Ratings.** FIFA attribute ratings from `players_stats` averaged across the active squad per team at match time, via `team_roster` stint dates. Contributes 88 features.

After joining all blocks, the working dataset contains `r nrow(df_joined)` matches before redundancy removal and cleaning.

The final dataset contains `r nrow(df_full)` matches across the 2024, 2025, and 2026 seasons after redundancy removal and cleaning. The reduction from `r nrow(matches)` raw matches arises from scraping coverage gaps in `match_team_stats` and early-season matches with no prior team rating snapshot.

## Redundancy Analysis

Before feature selection, all home/away column pairs were examined to identify statistically redundant features. For each pair, the mean sum (home + away) across all matches was computed. Table 3 shows the results, ordered by mean sum.

```{r pair-sums-table}
pair_sums %>%
  filter(!is.nan(mean_sum)) %>%
  filter(mean_sum >= 90 & mean_sum <= 110) %>%
  mutate(
    home_col = gsub("_", " ", home_col),
    mean_sum = round(mean_sum, 2),
    `Differential?` = ifelse(mean_sum >= 100, "Yes", "No")
  ) %>%
  select(`Feature (Home)` = home_col, `Mean Sum` = mean_sum, `Differential?`) %>%
  knitr::kable(caption = "Table 3: Home/away feature pairs with mean sums in the boundary region (90-110). Pairs with mean sum >= 100 are converted to differentials; those below are retained as independent features.")
```

Two categories of redundancy were identified. First, all possession band columns sum to exactly 100% across all matches — home and away possession are mathematically complementary. Second, passing accuracy and overall percentage statistics have mean sums above 100, measuring the same construct from both perspectives.

A threshold of **mean sum $\geq$ 100** was applied (line 1 in the code), cleanly separating these from independent count statistics such as shots, crosses, and fouls which sum well below 100 (line 2). All qualifying pairs were replaced with a single differential feature (home − away):

$$\Delta_j = x_{j,\text{home}} - x_{j,\text{away}}$$

A positive value indicates home team superiority. After this transformation, the feature space was reduced from `r ncol(df_joined) - 3` to `r ncol(df_full) - 3` features.

# Feature Selection

## Method 1 — Random Forest Variable Importance

A preliminary Random Forest model was trained on all `r ncol(X_all)` candidate features using 200 trees (`ntree = 200`) with balanced class weights. Variable importance was measured via the mean decrease in Gini impurity, which quantifies how much each feature contributes to homogeneous splits across all trees in the forest [@breiman2001]. Formally, the importance of feature $j$ is defined as:

$$I(j) = \frac{1}{B} \sum_{b=1}^{B} \sum_{t \in T_b} \Delta G(t, j)$$

where $B$ is the number of trees, $T_b$ is the set of nodes in tree $b$ that split on feature $j$, and $\Delta G(t, j)$ is the decrease in Gini impurity at node $t$. Figure 2 shows the top 15 features ranked by this criterion.

```{r rf-importance-plot, fig.cap="Figure 2: Top 15 features ranked by mean decrease in Gini impurity from the screening Random Forest.", fig.height=3.5}
imp_df %>%
  head(15) %>%
  mutate(
    feature = gsub("_", " ", feature),
    feature = reorder(feature, importance)
  ) %>%
  ggplot(aes(x = importance, y = feature)) +
  geom_col(fill = "steelblue") +
  labs(x = "Mean Decrease in Gini Impurity", y = "") +
  theme_bw(base_size = 10) +
  theme(axis.text.y = element_text(size = 9))
```

## Method 2 — Pearson Correlation with Target

The target variable was numerically encoded as $y \in \{1, 2, 3\}$ (away win = 1, draw = 2, home win = 3), and absolute Pearson correlation was computed for each feature:

$$r_j = \left| \frac{\sum_{i=1}^{n}(x_{ij} - \bar{x}_j)(y_i - \bar{y})}{\sqrt{\sum_{i=1}^{n}(x_{ij} - \bar{x}_j)^2 \sum_{i=1}^{n}(y_i - \bar{y})^2}} \right|$$

Using two independent methods — one tree-based and one linear — ensures that selected features have convergent evidence of predictive value. Figure 3 shows the top 15 features by absolute correlation.

```{r corr-importance-plot, fig.cap="Figure 3: Top 15 features ranked by absolute Pearson correlation with the encoded match outcome.", fig.height=3.5}
corr_df %>%
  head(15) %>%
  mutate(
    feature = gsub("_", " ", feature),
    feature = reorder(feature, correlation)
  ) %>%
  ggplot(aes(x = correlation, y = feature)) +
  geom_col(fill = "darkorange") +
  labs(x = "Absolute Pearson Correlation", y = "") +
  theme_bw(base_size = 10) +
  theme(axis.text.y = element_text(size = 9))
```

## Final Feature Set

Features appearing in the top 15 of both methods were retained, reducing the risk of including features predictive by chance in only one method. Table 4 lists the `r length(final_features)` selected features. Notably, FIFA-style squad and team rating features (Blocks C and D) did not appear in the top features of either method, suggesting in-match performance metrics are substantially more predictive than pre-match squad quality [@baboota2019].

```{r final-features-table}
feature_desc <- data.frame(
  Feature     = final_features,
  Description = gsub("_", " ", final_features)
)

knitr::kable(feature_desc,
  caption = "Table 4: Final selected features.",
  row.names = FALSE
)
```

# Model Development

## Temporal Train/Test Split

A temporal split was used rather than random partitioning — trained on 2024 ($n =$ `r nrow(train_df)`) and evaluated on 2025 ($n =$ `r nrow(test_df)`). This reflects real deployment conditions and avoids the optimistic bias of random splitting when observations are temporally autocorrelated [@baboota2019]. Table 5 confirms the class distribution is consistent across both sets.

```{r split-table}
split_summary <- data.frame(
  Set         = c("Training (2024)", "Test (2025)"),
  Matches     = c(nrow(train_df), nrow(test_df)),
  `Home Wins` = c(sum(train_df$result == "home_win"), sum(test_df$result == "home_win")),
  `Away Wins` = c(sum(train_df$result == "away_win"), sum(test_df$result == "away_win")),
  Draws       = c(sum(train_df$result == "draw"), sum(test_df$result == "draw"))
)

knitr::kable(split_summary, caption = "Table 5: Class distribution across training and test sets.")
```

## Random Forest Classifier

A Random Forest classifier [@breiman2001] was selected as it handles non-linear feature interactions, is robust to correlation, provides interpretable importance scores, and is less prone to overfitting than single decision trees through ensemble averaging.

The Random Forest model consists of $B$ classification trees, each trained on a bootstrap sample of the training data. At each split, only $m$ randomly selected features are considered, where $m$ is set to $\lfloor \sqrt{p} \rfloor$ and $p$ is the total number of features — the standard recommendation for classification tasks [@breiman2001]. The final prediction is the majority class across all trees:

$$\hat{y} = \text{mode}\left\{ h_b(\mathbf{x}) \right\}_{b=1}^{B}$$

where $h_b(\mathbf{x})$ is the prediction of the $b$-th tree for feature vector $\mathbf{x}$.

The model was trained with `ntree = 500` trees and `mtry =` `r floor(sqrt(length(final_features)))` features considered at each split. Class weights were set to equal values across all three outcome classes to prevent the model from being biased toward the majority class (home win).

# Results

## Evaluation on 2025 Test Data

The model was evaluated on the held-out 2025 season using overall accuracy, sensitivity, specificity, and macro-averaged F1-score. These metrics are defined as follows:

$$\text{Accuracy} = \frac{TP + TN}{TP + TN + FP + FN}$$

$$\text{Precision} = \frac{TP}{TP + FP}, \quad \text{Recall} = \frac{TP}{TP + FN}$$

$$F_1 = \frac{2 \cdot \text{Precision} \cdot \text{Recall}}{\text{Precision} + \text{Recall}}$$

The model achieved a test accuracy of **`r round(acc * 100, 1)`%** on the 2025 season, substantially above the no-information rate of `r round(max(table(y_test)) / length(y_test) * 100, 1)`% (the accuracy of always predicting the majority class). Table 6 presents the per-class performance metrics.

```{r performance-table}
perf <- as.data.frame(cm$byClass) %>%
  select(Sensitivity, Specificity, `Pos Pred Value`, `F1`) %>%
  round(3)
rownames(perf) <- c("Away Win", "Draw", "Home Win")
perf$Class <- rownames(perf)
perf <- perf %>% select(Class, Sensitivity, Specificity, `Pos Pred Value`, F1)

knitr::kable(perf,
  col.names = c("Class", "Sensitivity", "Specificity", "Precision", "F1-Score"),
  caption = paste0(
    "Table 6: Per-class performance metrics on the 2025 test set (overall accuracy = ",
    round(acc * 100, 1), "%)."
  )
)
```

```{r confusion-matrix-plot, fig.cap="Figure 4: Confusion matrix for 2025 test set predictions."}
cm_df <- as.data.frame(cm$table) %>%
  mutate(
    Prediction = recode(Prediction,
      home_win = "Home Win", away_win = "Away Win", draw = "Draw"
    ),
    Reference = recode(Reference,
      home_win = "Home Win", away_win = "Away Win", draw = "Draw"
    )
  )

ggplot(cm_df, aes(x = Reference, y = Prediction, fill = Freq)) +
  geom_tile(colour = "white", linewidth = 0.8) +
  geom_text(aes(label = Freq), size = 5, fontface = "bold") +
  scale_fill_gradient(low = "white", high = "steelblue") +
  labs(x = "Actual", y = "Predicted", fill = "Count") +
  theme_bw(base_size = 11) +
  theme(axis.text.x = element_text(angle = 15, hjust = 1, size = 10))
```

The model performs best at predicting home wins, which is consistent with the higher frequency of this class in the training data. Draw prediction is weakest, which is expected given the inherently lower predictability of drawn matches in football — a finding consistent with the broader sports prediction literature [@dixon1997].

xG and shots on goal differentials are the strongest predictors of match outcome, followed by late-game possession differentials and total crosses — consistent with the feature importance rankings from the screening stage.

## 2026 Validation

```{r results-2026-section, results='asis'}
if (has_2026) {
  cat(paste0(
    "To provide an additional out-of-sample validation, the model was retrained on all available ",
    "2024 and 2025 data ($n =$ ", nrow(train_full_df), " matches) and applied to the ",
    nrow(predict_2026), " MLS matches played in early 2026, for which results are already known.\n\n",
    "The retrained model achieved a validation accuracy of **", round(acc_2026 * 100, 1),
    "%** on the 2026 matches, compared to the 2025 test accuracy of ",
    round(acc * 100, 1), "%. This suggests the model generalises well to a new, unseen season ",
    "and that the selected features remain predictive beyond the training period."
  ))
} else {
  cat("The 2026 validation section will be populated once match statistics for the 2026 season have been scraped and uploaded to the database.")
}
```

```{r confusion-2026-plot, fig.cap="Figure 5: Confusion matrix for 2026 validation predictions.", eval=has_2026}
cm_2026_df <- as.data.frame(cm_2026$table) %>%
  mutate(
    Prediction = recode(Prediction,
      home_win = "Home Win", away_win = "Away Win", draw = "Draw"
    ),
    Reference = recode(Reference,
      home_win = "Home Win", away_win = "Away Win", draw = "Draw"
    )
  )

ggplot(cm_2026_df, aes(x = Reference, y = Prediction, fill = Freq)) +
  geom_tile(colour = "white", linewidth = 0.8) +
  geom_text(aes(label = Freq), size = 5, fontface = "bold") +
  scale_fill_gradient(low = "white", high = "steelblue") +
  labs(x = "Actual", y = "Predicted", fill = "Count") +
  theme_bw(base_size = 11) +
  theme(axis.text.x = element_text(angle = 15, hjust = 1, size = 10))
```

# Summary and Discussion

```{r summary-table}
summary_df <- data.frame(
  Stage = c(
    "Total matches in database", "Matches after cleaning",
    "Features before redundancy removal", "Features after redundancy removal",
    "Training set (2024)", "Test set (2025)",
    "Final features selected",
    "Test accuracy (2025)", "2026 validation accuracy"
  ),
  Result = c(
    format(nrow(matches), big.mark = ","),
    format(nrow(df_full), big.mark = ","),
    as.character(ncol(df_joined) - 3),
    as.character(ncol(df_full) - 3),
    format(nrow(train_df), big.mark = ","),
    format(nrow(test_df), big.mark = ","),
    as.character(length(final_features)),
    paste0(round(acc * 100, 1), "%"),
    ifelse(has_2026, paste0(round(acc_2026 * 100, 1), "%"), "Pending")
  )
)

knitr::kable(summary_df,
  col.names = c("Stage", "Result"),
  caption = "Table 7: Summary of dataset, modelling pipeline, and results."
)
```

This report developed a Random Forest classifier to predict MLS match outcomes using in-match team statistics, player statistics, and FIFA-style ratings. A systematic redundancy analysis converted complementary home/away feature pairs to differentials prior to model training. The model achieved a test accuracy of `r round(acc * 100, 1)`% on the held-out 2025 season and `r if(has_2026) paste0(round(acc_2026 * 100, 1), "% on early 2026 matches") else "a pending 2026 validation"`, substantially above the no-information baseline.

Limitations include reliance on post-match statistics (limiting pre-match forecasting utility), weak draw prediction, limited training data coverage, and residual class imbalance. Future work could explore rolling form features, pre-match betting odds, and gradient boosting methods such as XGBoost.


# References

```{r, include=FALSE}
# references rendered via bibliography field in YAML
```
